---
title: "Australian animals"
description: |
 Which Australian suburbs show most seasonality in dog noise complaints?
author:
  - name: Martine Wauben 
    url: https://github.com/MHWauben
date: 2020-07-25
categories:
  - RStats
  - Tidytuesday
  - Visualisation
preview: complaint-places.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(magrittr)
library(readr)
library(dplyr)
library(ggplot2)
library(rsample)
library(recipes)
library(parsnip)
library(randomForest)
```

## Tidy Tuesday: 21 July 2020

This week's data can be [found here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-07-28/readme.md). It is about penguin observations at the Palmer station. 

```{r load data, message = FALSE}
raw_penguins <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-28/penguins_raw.csv') %>%
  janitor::clean_names()
head(raw_penguins)
```

First: what's happening with the penguin species and islands? Do the species mix, or do they each have their own islands?

```{r vis islands}
ggplot(raw_penguins, aes(x = island, y = species))+
  geom_bin2d()+
  labs(title = 'Penguin species distribution on the Anvers islands',
       subtitle = 'Adelie penguins move about; the others stay put')+
  theme_minimal()
```

There are columns for egg shell isotopes. They are used as proxies for penguin health. However, can we use the same values for each species?

```{r vis eggs}
ggplot(raw_penguins, aes(x = delta_15_n_o_oo, y = delta_13_c_o_oo, colour = species))+
  geom_point()+
  labs(title = 'Eggshell isotopes of Palmer penguins',
       x = 'Delta 15 N o/oo',
       y = 'Delta 13 C o/oo',
       colour = '')+
  theme_minimal()+
  theme(legend.position = 'bottom')+
  guides(color = guide_legend(nrow = 2))
```

This dataset has missing values for several columns. Can we use ML to fill in the delta values with plausible values, based on measurements from the penguin's body? 

We'll start with the N value. 

```{r prep model data}
model_penguins <- raw_penguins %>%
  dplyr::select(body_mass_g, tidyselect::matches('_mm'), tidyselect::matches('oo')) %>%
  dplyr::filter(!is.na(body_mass_g))
true_missing <- dplyr::filter(model_penguins, is.na(delta_15_n_o_oo))


non_missing <- dplyr::filter(model_penguins, !is.na(delta_15_n_o_oo)) %>%
  dplyr::select(-delta_13_c_o_oo)
```

First, we split data into training and test sets: we will train the model using the training data, and then see how it performs on the testing data (which the model has not seen before).

```{r train test}
penguin_split <- rsample::initial_split(non_missing, prop = 0.8)
```

```{r prep recipe}
penguin_recipe <- rsample::training(penguin_split) %>%
  recipes::recipe(delta_15_n_o_oo ~ .) %>%
  recipes::step_center(recipes::all_predictors(), 
                       -recipes::all_outcomes()) %>%
  recipes::step_scale(recipes::all_predictors(), 
                      -recipes::all_outcomes()) %>%
  recipes::prep()

penguin_test <- penguin_recipe %>%
  recipes::bake(rsample::testing(penguin_split))
```

These steps pre-process the data, so it's nicely centered and scaled so outliers should not affect our results as much. To see what our training looks like, we use the `juice` function.

```{r glimpse prep training data}
penguin_training <- recipes::juice(penguin_recipe)
head(penguin_training)
```

```{r init models}
pen_rf <- parsnip::rand_forest(trees = 100, mode = "regression") %>%
  parsnip::set_engine("randomForest") %>%
  parsnip::fit(delta_15_n_o_oo ~ ., data = penguin_training)
```

```{r predict outcomes}
penguin_test$predictions <- predict(pen_rf, penguin_test)[[1]]
ggplot(penguin_test, aes(x = delta_15_n_o_oo, y = predictions))+
  geom_point()+
  geom_abline(intercept = 0, slope = 1, colour = 'blue')+
  coord_fixed()+
  xlim(c(8,10))+
  ylim(c(8, 10))
```

```{r accuracy metrics}
yardstick::metrics(penguin_test, truth = delta_15_n_o_oo, estimate = predictions)
```

